---
title: "Introdução à estatística"
output: html_document
date: "2024-11-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r fig.height=2, fig.width=5, dpi = 72}
library(tidyverse)
library(esquisse)
library(patchwork)
library(here)
library(ggsci)
```

#Um grupo
```{r fig.height=8, fig.width=5, dpi = 72}
set.seed = 123

normal_dist = rnorm(200, mean = 0, sd = 1) %>% 
  as.data.frame() %>% 
  rename(distribution = ".") %>% 
  mutate(group = "G1")

normal_dist_params = normal_dist %>% 
  summarize(mean = mean(distribution),
            median = median(distribution))

right_skew = rchisq(100, df = 3) %>% 
  as.data.frame() %>% 
  rename(distribution = ".") %>% 
  mutate(group = "G1")

right_skew_params = right_skew %>% 
  summarize(mean = mean(distribution),
            median = median(distribution))

left_skew = -sqrt(right_skew$distribution) %>% 
  as.data.frame() %>% 
  rename(distribution = ".") %>% 
  mutate(group = "G1")

left_skew_params = left_skew %>% 
  summarize(mean = mean(distribution),
            median = median(distribution))
```

Normal
```{r fig.height=5, fig.width=5, dpi = 72}
normal_points = normal_dist %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_point(size = 2.5,
             color = "black") +
  geom_point(size = 2,
             color = "white") +
  geom_vline(xintercept = normal_dist_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = normal_dist_params$median, color = "#05C7F2FF",
             size = 1) +
  theme_minimal() +
      scale_x_continuous(expand = c(0.01,0.01))+

  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12)) +
  labs(y = "",
       x = "")


normal_jitter = normal_dist %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_jitter(shape = 21, size = 2.5, stroke = 0.5, color = "black", fill = "white") +
    geom_vline(xintercept = normal_dist_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = normal_dist_params$median, color = "#05C7F2FF",
             size = 1) +
    scale_x_continuous(expand = c(0.01,0.01))+
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12)) +
  labs(y = "",
       x = "")



normal_hist = normal_dist %>% 
  ggplot() +
  aes(x = distribution) +
  geom_histogram(color = "white",
                 fill = "#1D373AFF",
                 size = 0.5,
                 alpha = 1,
                 bins = 10) +
    geom_vline(xintercept = normal_dist_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = normal_dist_params$median, color = "#05C7F2FF",
             size = 1) +
  scale_x_continuous(expand = c(0.01, 0.01))+
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12),
        panel.grid.minor.y = element_blank()) +
  labs(y = "",
       x = "")

normal_boxplot = normal_dist %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_boxplot() +
    geom_vline(xintercept = normal_dist_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = normal_dist_params$median, color = "#05C7F2FF",
             size = 1) +
    scale_x_continuous(expand = c(0.01,0.01))+
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12)) +
  labs(y = "",
       x = "")

normal_violin = normal_dist %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_violin(fill = "black", alpha = 0.1) +
  geom_jitter(shape = 21, size = 2.5, stroke = 0.5, color = "black", fill = "white",
              width = 0.0005) +
  geom_boxplot(width = 0.3) +
    geom_vline(xintercept = normal_dist_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = normal_dist_params$median, color = "#05C7F2FF",
             size = 1) +
    scale_x_continuous(expand = c(0.01,0.01))+
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12)) +
  labs(y = "",
       x = "")


patch_normal_dist = normal_points /
  normal_jitter/
  normal_hist /
  normal_boxplot +
  normal_violin + 
  plot_layout(axes = "collect",
              heights = c(0.2, 0.5, 1, 1, 1))

patch_normal_dist

patch_normal_dist %>% 
  ggsave(filename = here("Figuras", "Distribuicao_normal.png"),
         width = 5,
         height = 8)
```


Deslocado para a direita
```{r fig.height=5, fig.width=5, dpi = 72}
right_skew_points = right_skew %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_point(size = 2.5,
             color = "black") +
  geom_point(size = 2,
             color = "white") +
  geom_vline(xintercept = right_skew_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = right_skew_params$median, color = "#05C7F2FF",
             size = 1) +
  theme_minimal() +
      scale_x_continuous(expand = c(0.01,0.01), n.breaks = 20)+

  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12)) +
  labs(y = "",
       x = "")


right_skew_jitter = right_skew %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_jitter(shape = 21, size = 2.5, stroke = 0.5, color = "black", fill = "white") +
    geom_vline(xintercept = right_skew_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = right_skew_params$median, color = "#05C7F2FF",
             size = 1) +
      scale_x_continuous(expand = c(0.01,0.01), n.breaks = 20)+
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12)) +
  labs(y = "",
       x = "")



right_skew_hist = right_skew %>% 
  ggplot() +
  aes(x = distribution) +
  geom_histogram(color = "white",
                 fill = "#1D373AFF",
                 size = 0.5,
                 alpha = 1,
                 bins = 10) +
    geom_vline(xintercept = right_skew_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = right_skew_params$median, color = "#05C7F2FF",
             size = 1) +
      scale_x_continuous(expand = c(0.01,0.01), n.breaks = 20)+
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12),
        panel.grid.minor.y = element_blank()) +
  labs(y = "",
       x = "")

right_skew_boxplot = right_skew %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_boxplot() +
    geom_vline(xintercept = right_skew_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = right_skew_params$median, color = "#05C7F2FF",
             size = 1) +
      scale_x_continuous(expand = c(0.01,0.01), n.breaks = 20)+
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12)) +
  labs(y = "",
       x = "")

right_skew_violin = right_skew %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_violin(fill = "black", alpha = 0.1) +
  geom_jitter(shape = 21, size = 2.5, stroke = 0.5, color = "black", fill = "white",
              width = 0.0005) +
  geom_boxplot(width = 0.3) +
    geom_vline(xintercept = right_skew_params$mean, color = "black",
             size = 1,) +
  geom_vline(xintercept = right_skew_params$median, color = "#05C7F2FF",
             size = 1) +
      scale_x_continuous(expand = c(0.01,0.01), n.breaks = 20)+
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12)) +
  labs(y = "",
       x = "")


patch_right_skew_dist = right_skew_points /
  right_skew_jitter/
  right_skew_hist /
  right_skew_boxplot +
  right_skew_violin + 
  plot_layout(axes = "collect",
              heights = c(0.2, 0.5, 1, 1, 1))

patch_right_skew_dist

patch_right_skew_dist %>% 
  ggsave(filename = here("Figuras", "Distribuicao_right_skewed.png"),
         width = 5,
         height = 8)
```



# Dois grups

Dois grupos, pvalue < 0.00000005
```{r fig.height=5, fig.width=5, dpi = 72}

set.seed = 123

normal_dist_1 = rnorm(100, mean = 30, sd = 2) %>% 
  as.data.frame() %>% 
  rename(distribution = ".") %>% 
  mutate(group = "G1")

normal_dist_2 = rnorm(100, mean = 50, sd = 2) %>% 
  as.data.frame() %>% 
  rename(distribution = ".") %>% 
  mutate(group = "G2")

normal_dist_2groups = bind_rows(normal_dist_1, 
                                normal_dist_2)

normal_dist_2groups_params = normal_dist_2groups %>% 
  group_by(group) %>% 
  summarize(mean = mean(distribution),
            median = median(distribution)) %>% 
  ungroup()

normal_points = normal_dist_2groups %>% 
  ggplot() +
  aes(x = distribution, 
      y = group,
      color = group) +
  geom_point(size = 2.5,
             color = "black") +
  geom_point(size = 2) +
  theme_minimal() +
  scale_fill_manual(values = c("black", "#05C7F2FF")) +
  scale_color_manual(values = c("black", "#05C7F2FF")) +
  scale_x_continuous(expand = c(0.1,0.1), n.breaks = 10)+

  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12),
        legend.position = "none") +
  labs(y = "",
       x = "")

normal_violin = normal_dist_2groups %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_violin(aes(fill = group),
              alpha = 0.1) +
  geom_boxplot(aes(fill = group),width = 0.3) +
  scale_x_continuous(expand = c(0.1,0.1), n.breaks = 10)+
  scale_fill_manual(values = c("black", "#05C7F2FF")) +
  scale_color_manual(values = c("black", "#05C7F2FF")) +
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12),
        legend.position = "none") +
  labs(y = "",
       x = "")


normal_hist = normal_dist_2groups %>% 
  ggplot() +
  aes(x = distribution, 
      fill = group) +
   geom_histogram(bins = 20L, alpha = 0.5) +
  scale_x_continuous(expand = c(0.1,0.1), n.breaks = 10)+
  scale_fill_manual(values = c("black", "#05C7F2FF")) +
  scale_color_manual(values = c("black", "#05C7F2FF")) +
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12),
        legend.position = "none") +
  labs(y = "",
       x = "")


patch_normal_dist_2groups = normal_points /
  normal_violin/
  normal_hist + 
  plot_layout(axes = "collect",
              guides = "collect",
              heights = c(0.5, 1, 1))

patch_normal_dist_2groups

patch_normal_dist_2groups %>% 
  ggsave(filename = here("Figuras", "Distribuicao_2grupos.png"),
         width = 5,
         height = 3)
```

2 grupos, pouco separados
```{r fig.height=5, fig.width=5, dpi = 72}

set.seed = 123

normal_dist_1 = rnorm(100, mean = 45, sd = 1) %>% 
  as.data.frame() %>% 
  rename(distribution = ".") %>% 
  mutate(group = "G1")

normal_dist_2 = rnorm(100, mean = 47, sd = 1) %>% 
  as.data.frame() %>% 
  rename(distribution = ".") %>% 
  mutate(group = "G2")

normal_dist_2groups = bind_rows(normal_dist_1, 
                                normal_dist_2)

normal_dist_2groups_params = normal_dist_2groups %>% 
  group_by(group) %>% 
  summarize(mean = mean(distribution),
            median = median(distribution)) %>% 
  ungroup()

normal_points = normal_dist_2groups %>% 
  ggplot() +
  aes(x = distribution, 
      y = group,
      color = group) +
  geom_point(size = 2.5,
             color = "black") +
  geom_point(size = 2) +
  theme_minimal() +
  scale_fill_manual(values = c("black", "#05C7F2FF")) +
  scale_color_manual(values = c("black", "#05C7F2FF")) +
  scale_x_continuous(expand = c(0.1,0.1), n.breaks = 10)+

  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12),
        legend.position = "none") +
  labs(y = "",
       x = "")

normal_violin = normal_dist_2groups %>% 
  ggplot() +
  aes(x = distribution, 
      y = group) +
  geom_violin(aes(fill = group),
              alpha = 0.1) +
  geom_boxplot(aes(fill = group),width = 0.3) +
  scale_x_continuous(expand = c(0.1,0.1), n.breaks = 10)+
  scale_fill_manual(values = c("black", "#05C7F2FF")) +
  scale_color_manual(values = c("black", "#05C7F2FF")) +

  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12),
        legend.position = "none") +
  labs(y = "",
       x = "")


normal_hist = normal_dist_2groups %>% 
  ggplot() +
  aes(x = distribution, 
      fill = group) +
   geom_histogram(bins = 20L, alpha = 0.5) +
  scale_x_continuous(expand = c(0.1,0.1), n.breaks = 10)+
  scale_fill_manual(values = c("black", "#05C7F2FF")) +
  scale_color_manual(values = c("black", "#05C7F2FF")) +
  theme_minimal() +
  theme(axis.line.x = element_line(size = 1),
        text = element_text(color = "black",
                            size = 12),
        legend.position = "none") +
  labs(y = "",
       x = "")


patch_normal_dist_2groups = normal_points /
  normal_violin/
  normal_hist + 
  plot_layout(axes = "collect",
              guides = "collect",
              heights = c(0.5, 1, 1))

patch_normal_dist_2groups

patch_normal_dist_2groups %>% 
  ggsave(filename = here("Figuras", "Distribuicao_2grupos_poucosep.png"),
         width = 5,
         height = 3)
```

